# Project

# Introduction
In this analysis we will explore gene expression analysis among TNBC and ER+ genes using Bioconductor packages.

Load the packages [research on where the packages came from]

```{r}

library(omicsdata)
library(limma)
library(EDASeq)
library(edgeR)
library(dplyr)
library(plyr)
library(corrplot)
library(ggfortify)
library(ggplot2)
```


## Exploratory analysis

```{r}
breastcancer <- readRDS("/opt/omicsdata/datasets/SRP042620.rds") #load the data
breastcancer #visualize the data
```

```{r}
class(breastcancer) # to check the class of our data
```

```{r}
assay(breastcancer) #we notice the first five genes have zero count, hence we will filter out genes with zero counts
```


```{r}
#we use the function below to identify genes that have all zero counts across all samples and remove them from the dataset.
table(rowSums(assay(breastcancer, "raw_counts")) == 0) 


```

We have 5889 genes of zero counts. Now we will filter out these genes

```{r}
breastcancer <- breastcancer[rowSums(assay(breastcancer, "raw_counts")) > 0, ] # We keep the genes with the sum > 0
breastcancer
```
At the beginning of this analysis the unfiltered dataset had  63856 genes and 168 samples.
After filtering the genes with zero counts we see that filtered dataset has now 57967 genes and 168 samples. hence, we correctly excluded 5889 genes with zero counts.

#Biological Question: 

The aim of our analysis is to identify the transcriptional differences between the two cancer types: Triple Negative Breast Cancer and ER+ Primary Cancer. The subtype information is stored in the $sra.sample_attributes in colData() object.

```{r}
table(breastcancer$sra.sample_attributes)
```
We have got 28 Cancer Cell Line|tissue, 42 - ER+, 42 - Triple Negative Breast Cancer, 5 No known cancer, 30 -Tissue adjacent to ER+ Primary Tumor, 21 - Tissue adjacent to TNBC Primary Tumor.
The scope of this analysis is to learn  which genes are different between TNBC and ER+. Hence, we will filter out other cancer types to keep: TNBC and ER+.


```{r}
#create groups
breastcancer$group <- NA
breastcancer$group[grepl("ER\\+ Breast Cancer Primary Tumor", breastcancer$sra.sample_attributes)] <- "ER" # ER (ONLY primary tumor)
breastcancer$group[grepl("Triple Negative Breast Cancer Primary Tumor", breastcancer$sra.sample_attributes)] <- "TNBC" # TNBC (ONLY primary tumor)
```

```{r}
table(breastcancer$group) #to get the counts for each group
```


```{r}
breastcancer$group # to visualize the column of group
```
In summary, We created a grouping variable in the sample metadata to label ER and TNBC samples.

```{r}
breastcancer <- breastcancer[, !is.na(breastcancer$group)] #filter the dataset
```

Second we subset the SummarizedExperiment to retain only those samples, optionally overwriting the original object to focus the analysis on the comparison of interest.

```{r}
#confirm that there are no NA in our dataset
sum(is.na(breastcancer$group))
summary(breastcancer$group)
```

```{r}
# We visualize the first six genes and the first six samples
assay(breastcancer, "raw_counts")[1:6, 1:6]
```

## boxplot

```{r}
raw_counts <- assay(breastcancer, "raw_counts")
boxplot(raw_counts[, 1:20])
```
```{r}
boxplot(log1p(raw_counts)[, 1:20]) #apply log1p transformation then plot
```

```{r}
colData(breastcancer)
```

```{r}
library(tidyverse)

# 1. Extract counts
raw_counts <- assay(breastcancer, "raw_counts")

# 2. Shorten sample names (same as base R)
colnames(raw_counts) <- substr(colnames(raw_counts), 9, 20)

# 3. Log transform + convert to data frame
df <- as.data.frame(log1p(raw_counts))

# 4. Keep gene info
df$gene <- rownames(df)

# 5. Convert to long format
df_long <- df %>%
  pivot_longer(cols = -gene,
               names_to = "sample",
               values_to = "expression")

df_long

# 6. Add subtype (biological group)
df_long$subtype <- breastcancer$group[
  match(df_long$sample, colnames(raw_counts))
]

df_long

# 7. Keep only first 20 samples (same as original)
samples_84 <- colnames(raw_counts)[1:84]
df_long <- df_long %>%
  filter(sample %in% samples_84)

# 8. Plot
ggplot(df_long, aes(x = sample, y = expression, fill = subtype)) +
  geom_boxplot() +
  scale_y_continuous(limits = c(0, 20)) +
  labs(x = "Sample", y = "log1p expression", fill = "subtype") +
  theme(axis.text.x = element_text(angle = 90, size = 7))
```

## MD and MA charts
```{r}
# comparing gene vs itself across samples (Y-axis) and positioning it by overall expression level (X-axis)
limma::plotMA(log1p(raw_counts), main = "", xlab = "A", ylab = "M")
```

X axis - Genes near 12 are highly expressed whereas genes near 0 are lowly expressed.
Y axis - Genes near +5 show much higher expression than average. Whereas genes near -5 show much lower expression than average.
Dots on the top right (high A and high M): highly expressed genes that are even higher than average
Dots on the lower right (high A and low M): highly expressed genes that are lower than average
Dots on the left side: lowly expressed genes.

Ideally, we would expect a horizontal line right at the 0 y-axis. Indeed, genes have different expression levels which are spread on the X-axis but should have no systematic bias (shown on the Y axis).

The MA plot compares each gene to itself across samples, not to other genes. The x-axis (A) represents how strongly a gene is expressed overall, effectively placing it relative to other genes in terms of expression level. The y-axis (M) represents how much that same gene differs in a specific sample compared to its average expression across all samples. This means a gene can be highly expressed overall (high A) but still show a negative M value if it is expressed less than its usual level in that particular sample.



While, to visualize the MD plot we can use the MDplot() function of the *EDASeq* package. In particular, we choose a pair of experiments for comparison (6 and 7 samples). In this case, we use the raw count matrix.

The MD plot is useful because it directly compares gene expression between two specific samples, allowing detection of pairwise differences and biases that may not be visible when comparing a sample to the overall average. While the MA plot assesses global bias relative to all samples, the MD plot helps identify inconsistencies, batch effects, or outlier samples by focusing on direct sample-to-sample variation.


```{r}
MDPlot(raw_counts, c(6, 7))
```

Even though samples 6 and 7 belong to the same cancer type, most genes show similar expression (centered around 0), indicating overall consistency. However, the presence of a systematic trend (non-flat line) suggests technical bias between the samples, especially depending on expression level. Additionally, higher variability is observed in low-expression genes, and some outliers indicate strong differences. This suggests that normalization is needed before further analysis.


## RLE plot

RLE plots are useful to check the differences between samples. Ideally, the median of boxplot should be centered around zero. Deviations indicate technical variation or differences in library size that may need normalization.

much it deviates from the typical value.

```{r}
#create colors for RLE graph plot
col <- as.factor(breastcancer$group)
levels(col)
levels(col) <- c(1:2) 

plotRLE(raw_counts, outline = FALSE, las = 2,
        ylab = "RLE", main = "RLE of raw counts", col = col)

```

NOTE: [CHNAGE NAME OF SAMPLES] BECAUSE: The samples were not reordered. The RLE plot uses the column names of the raw count matrix as labels on the x-axis. Since the sample names were shortened using substr(), they now appear as partial IDs instead of sequential numbers, which makes them look unordered. However, the actual order of samples in the dataset has not changed.

## PCA
[why does the pca looks the same to what we did in the class???]

```{r}
raw_counts <- assay(breastcancer, "raw_counts")

# log transform
pca <- prcomp(t(log1p(raw_counts)))

# make sure group is a factor
colData(breastcancer)$group <- factor(colData(breastcancer)$group, levels = c("ER", "TNBC"))

# PCA plot colored by cancer type
autoplot(
  pca,
  data = as.data.frame(colData(breastcancer)),
  colour = "group",
  size = 3
) +
  labs(title = "PCA of TNBC and ER samples", colour = "Cancer type") +
  theme_bw()


```

```{r}
#To label samples:
autoplot(
  pca,
  data = as.data.frame(colData(breastcancer)),
  colour = "group",
  label = TRUE,
  label.size = 3,
  size = 3
) +
  theme_bw()
```
```{r}
screeplot(pca, type = c("lines"))
```

# GC content

GC content bias evaluation checks whether gene expression measurements are affected by the proportion of G (guanine) and C (cytosine) nucleotides in each gene. Ideally, gene expression should reflect true biological signal, but in practice, sequencing technologies can preferentially capture or amplify genes with certain GC content. This introduces a technical bias, meaning some genes may appear more or less expressed not because of biology, but because of their sequence composition.

To assess this, we first retrieve GC content and gene length for each gene using annotation databases. The code extracts gene IDs from the dataset and uses them to query external genome annotation (hg38). The function getGeneLengthAndGCContent() returns, for each gene, its length and GC percentage. These values are then used in later steps to check whether expression levels depend on GC content. Genes without GC or length information are removed because they cannot be used to evaluate this bias.

```{r}
rowRanges(breastcancer)
```

```{r, message=F}

ensid <- as.character(rowRanges(breastcancer)$gene_id) # Extract gene IDs from the dataset (metadata per gene)
ensid <- substr(ensid, 1, 15) #Keep only the core gene ID, remove version info
ensid

# we specify genome assembly hg38 and org.db as mode to retrieve the information
gc <- getGeneLengthAndGCContent(id = ensid, org = "hg38", mode = "org.db")
gc
# org = which genome / annotation reference to use
# mode = org.db meaning we are using a pre-built annotation database in R (fast and reliable)
```

We obtain a numeric matrix, labeled us "gc", with two columns: gene length and GC-content.
 we notice that some genes have no information (i.e., NA) about GC and gene length, so we to eliminate them and check for GC bias.

```{r}
table(is.na(gc[,2])) # look at column 2 of gc and count how many are NA (i.e, 31110 genes)

```

```{r}
gc <- gc[!is.na(gc[,2]),] # this function keeps only genes, where gc is not missing.
table(is.na(gc[,2]))
```
Now we can see that we removed the gens with NA.

```{r}
rowData(breastcancer)$ensid = ensid # adding gene IDs from ensid as column inside breastcancer. This is because breastcancer gene IDs were not easily usable. Now you can match genes with gc. [correct]

```

```{r}
breastcancer <- breastcancer[rowData(breastcancer)$ensid %in% rownames(gc),] #Compare genes and breastcancer vs gc to keep only genes found in gc(i.e., genes with valid GC info)
```

```{r}
raw_counts <- assay(breastcancer,  "raw_counts") # delete this command from here to above
row.names(raw_counts) <- row.names(gc) # this done so both objects (raw_counts" and gc) use the same gene IDs and can be compared.
```

```{r}
table(duplicated(row.names(raw_counts)))
```
We can see that we don't have duplicates.

Then we create an object of class **SeqExpressionSet** to visualize the GC content. in particular, we use `newSeqExpressionSet()` function of the *EDASeq* package.

```{r}
set <- newSeqExpressionSet(raw_counts,
                           phenoData = AnnotatedDataFrame(data.frame(conditions= factor(1:ncol(raw_counts)),
                                                                     row.names = colnames(raw_counts))),
                           featureData = AnnotatedDataFrame(data.frame(gc = gc[,2], l = gc[,1])))
set
```
We add a dummy variable because the SetExpressionset object requires sample-level metadata to be created, even if we are not using it for analysis at this stage. Since we are only interested in GC bias (a gene level effect), we assign simple placeholder values to satisfy the structure, allowing the functions to run without introducing biological meaning yet.



To visulaize the GC content for each sample we use the `biasPlot()` function.

```{r}
EDASeq::biasPlot(
  set,
  "gc",
  ylim = c(0, 15),
  log = TRUE,
  xlab = "GC content",
  ylab = "Log gene counts",
  legend = TRUE,
  col = 1:ncol(counts(set))
)

legend(
  "topleft",
  legend = colnames(counts(set)),
  col = 1:ncol(counts(set)),
  lty = 1,
  title = "Samples"
)
```

Interpretation of your plot

This is a GC-content bias plot:

X-axis (GC content): proportion of G/C bases in genes
Y-axis (log gene counts): expression level (log-scaled)

What your plot shows:

All curves have a similar shape → good consistency across samples
Counts increase with GC content up to ~0.6, then decline
This indicates a systematic GC bias (common in RNA-seq)

How to interpret it biologically/technically:

The peak around ~60% GC suggests sequencing or amplification favors mid-GC genes
The drop at high GC may reflect PCR or alignment inefficiencies
Since all samples behave similarly, the bias is technical and consistent, not sample-specific


Now, we will see within-samples normalization for GC-content bias and the between-samples normalization.

```{r}
save(list = c("breastcancer", "raw_counts", "set", "gc"), file = "preprocessingG6.RData")
```






